---
title: "example ts modeling"
author: "Andrés Camargo"
date: "2023-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Data

```{r}
library(tidyverse)
library(lubridate)
library(Metrics)
library(zoo)
library(forecast)
library(prophet)
library(ggplot2)

s_ts = read.csv("data/sales_products_ts.csv")

s_ts = s_ts %>% select(c(id,date, quantity)) %>% mutate(date= ymd(date))

# Select the most sold product
# s_ts %>% group_by(id) %>% summarise(total_qty = sum(quantity)) %>% arrange(desc(total_qty)) # id 17410306

product = s_ts %>% filter(id=='17410306') %>% 
  mutate(date_ym = paste(year(date), month(date), sep = "-")) %>% 
  arrange(date)

product
```

Length of the time series

```{r}
unique(year(product$date))
```
Plot the time series:

```{r}
product %>% ggplot(aes(date, quantity)) + geom_line()
```

*Remove outliers*

Remove outliers identified (Q1 -+ 3 × IQR), and replaces them with linearly interpolated replacements.
```{r}
boxplot(product$quantity)
q3 = quantile(product$quantity, 0.75)
q1 = quantile(product$quantity, 0.25)
iqr = IQR(product$quantity, na.rm = TRUE)
maxLimit = q3 + (1.5*iqr)
minLimit = q1 - (1.5*iqr)

replace_outliers <- function(x, max, min) {
  x[x > max] <- max
  x[x < min] <- min
  return(x)
}

product$quantity = replace_outliers(product$quantity, maxLimit, minLimit)

# attempt with tsclean() function of forecast
```


Logaritmic transformation to reduce variance

```{r}
product['qtyLog'] = log(product$quantity+1)
product

# plot the log
product %>% ggplot(aes(date, qtyLog)) + geom_line()
```

## Train/test split

```{r}
library(TSstudio)

product = product %>% mutate(test=ifelse(date>='2021-01-01',1,0))
ggplot(product, aes(date, quantity, color=test))+geom_line()
train = product %>% filter(test== 0)
test = product %>% filter(test== 1)

```


## Baseline models

### 1.1 Constant model

```{r}
mean_model = mean(product$qtyLog)
product['qtyMean'] = exp(mean_model)-1

product
```
Plot constant model

```{r}
product %>% ggplot(aes(date)) + 
  geom_line(aes(y = quantity, color = "actual")) +
  geom_line(aes(y = qtyMean, color = "Mean")) +
  labs(y = "Quantity", color = "Series")
```

Compute RMSE *New Data Frame for predictions and errors*

```{r}
rmse_na = function (x, y) {
  sqrt(mean((x - y)^2, na.rm = TRUE))
}

results_new = data.frame(qty = product$quantity)
results_new$f_mean = exp(mean_model)-1
results_new$rmse_mean = rmse_na(results_new$qty, results_new$f_mean)

results_new
```


### 1.1 Linear Model

Create time difference

```{r}
min(product$date) # "2015-12-28"
product$timeIndex <- product$date - min(product$date)

product$timeIndex <- interval(floor_date(min(product$date), unit = "week"), floor_date(product$date, unit = "week")) / weeks(1)

product

```

```{r}
model_linear <- lm(qtyLog ~ timeIndex, data = product)
summary(model_linear)

model_linear_pred = predict(model_linear)

product$qtyLinear = exp(model_linear_pred)-1

product %>% ggplot(aes(date)) + 
  geom_line(aes(y = quantity, color = "actual")) +
  geom_line(aes(y = qtyLinear, color = "linear")) +
  geom_line(aes(y = qtyMean, color = "mean")) +
  labs(y = "Quantity", color = "Series")
```

Save RMSE

```{r}
results_new$f_linear = exp(model_linear_pred)-1
results_new$rmse_linear = rmse_na(results_new$qty, results_new$f_linear)
results_new
```


### 1.3 Random Walk

Create Yt-1: 

```{r}
product$qtyLogShift1 = lag(product$qtyLog)
product

```

Plot shift 1

```{r}
product %>% ggplot(aes(qtyLogShift1, qtyLog)) + geom_point()

```

Calculate and plot the difference

```{r}
product$qtyLogDiff = product$qtyLog - product$qtyLogShift1

product %>% ggplot(aes(date)) + 
  geom_line(aes(y = qtyLogDiff, color = "actual")) +
  labs(y = "Quantity", color = "Series")

```

```{r}
# create column with predictions of the random walk
product$qtyRandomW = exp(product$qtyLogShift1)-1

# plot random walk
product %>% ggplot(aes(date)) + 
  geom_line(aes(y = quantity, color = "actual")) +
  geom_line(aes(y = qtyRandomW, color = "Random Walk")) +
  labs(y = "Quantity", color = "Series")
```


Save results for random walk

```{r}
results_new$f_randomW = exp(product$qtyLogShift1)-1
results_new$rmse_randomW = rmse_na(results_new$qty, results_new$f_randomW)
results_new
```

### 2.1 Moving Average Model

```{r}
product$qtyLogMA4 = rollmean(product$qtyLog, k=4, allign="right", fill=NA) 

product$qtyMA4 = exp(product$qtyLogMA4)-1

product %>% ggplot(aes(date)) + 
  geom_line(aes(y = qtyLog, color = "actual")) +
  geom_line(aes(y = qtyLogMA4, color = "Mean Average 4")) +
  labs(y = "Quantity", color = "Series")

```


```{r}
product %>% ggplot(aes(date)) + 
  geom_line(aes(y = quantity, color = "Actual")) +
  geom_line(aes(y = qtyMean, color = "Mean")) +
  geom_line(aes(y = qtyLinear, color = "Linear")) +
  geom_line(aes(y = qtyRandomW, color = "Random W")) +
  geom_line(aes(y = qtyMA4, color = "Mean Average 4")) +
  labs(y = "Quantity", color = "Series")
```

results for moving average new DF
```{r}
results_new$f_moving_a4 = exp(product$qtyLogMA4)-1
results_new$rmse_ma4 = rmse_na(results_new$qty, results_new$f_moving_a4)
results_new
```

## 3. ARIMA

```{r}
ts = product$qtyLog
```

ACF and PACF for hyperparameters selection:

```{r}
lag_acf = acf(ts, lag.max = 52)
```

```{r}
lag_pacf = pacf(ts, lag.max = 52)
```

Model

```{r}
# new model
model_arima =auto.arima(ts)

summary(model_arima)
```

```{r}
# extract the results
fited= as.data.frame(fitted(model_arima))

product$qtyLogArima = fited$x
product$qtyArima = exp(product$qtyLogArima)-1

product %>% ggplot(aes(date)) + 
  geom_line(aes(y = quantity, color = "Actual")) +
  geom_line(aes(y = qtyArima, color = "Arima")) +
  labs(y = "Quantity", color = "Series")
```

```{r}
# estimate error
results_new$f_arima = product$qtyArima
results_new$rmse_arima = rmse_na(results_new$qty, results_new$f_arima)
results_new

```

## 4. Prophet

```{r}
ts_date = product %>% dplyr::select(c(date, qtyLog)) %>%
  rename(ds = date, y = qtyLog)

model_prophet = prophet(ts_date)

summary(model_prophet)
```

Fit and results

```{r}
fit_prophet = predict(model_prophet)
product$qtyLogProphet = fit_prophet$yhat
product$qtyProphet = exp(product$qtyLogProphet)-1

product

plot(model_prophet, fit_prophet)

prophet_plot_components(
  model_prophet,
  fit_prophet,
  uncertainty = TRUE,
  plot_cap = TRUE,
  weekly_start = 0,
  yearly_start = 0,
  render_plot = TRUE
)

results_new$f_prophet = product$qtyProphet
results_new$rmse_prophet = rmse_na(results_new$qty, results_new$f_prophet)
results_new


#write.csv(results_new, "results/results_1st_base.csv")
```

```{r}
product %>% ggplot(aes(date)) + 
  geom_line(aes(y = quantity, color = "Actual")) +
  #geom_line(aes(y = qtyArima, color = "Arima")) +
  geom_line(aes(y = qtyProphet, color = "Prophet")) +
  labs(y = "Quantity", color = "Series") + theme_minimal()
```

Cross Validation

```{r}
# train up to 2021
ts_cv_d <- cross_validation(model_prophet, initial = 1807, period = 90, horizon = 30, units = 'days')

ts_cv_d

metrics_prophet= performance_metrics(ts_cv_d)
metrics_prophet


```





Forecast

```{r}
future = make_future_dataframe(model_prophet, periods= 8)
forecast = predict(model_prophet, future)
tail(forecast[c('ds', 'yhat', 'yhat_lower', 'yhat_upper')])

forecast %>% ggplot(aes(ds, yhat)) + geom_line()
plot(model_prophet, forecast)
```


